% !TEX root = thesis.tex
\chapter{Towards Efficient Inference: Techniques and Solutions}\label{chap:methods}

Drawing on the bottleneck analysis of Chapter~\ref{chap:inference}, this chapter systematically surveys the techniques that have been developed to reduce the resource demands of \ac{llm} inference. For each technique the discussion covers its core mechanism, the bottleneck it addresses, its accuracy--efficiency trade-off, and its practical implications for edge deployment. Figure~\ref{fig:methods_map} provides an overview of how the techniques presented in this chapter map to the bottlenecks identified in Chapter~\ref{chap:inference}.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
  node distance   = 0.35cm and 0.5cm,
  bottleneck/.style = {rectangle, draw=red!70, fill=red!8, text width=4.5cm,
                       text centered, rounded corners=3pt,
                       minimum height=1.6em, font=\small\bfseries},
  solution/.style   = {rectangle, draw=blue!70, fill=blue!8, text width=4.3cm,
                       text centered, rounded corners=3pt,
                       minimum height=1.4em, font=\small},
  header/.style     = {font=\small\bfseries, align=center},
  arrow/.style      = {-{Stealth}, thick, red!60}
]
%% Headers
\node [header] (h1) at (0,0)   {Bottleneck};
\node [header] (h2) at (7.5,0) {Technique};

%% Bottleneck nodes
\node [bottleneck, below=0.5cm of h1] (b1) {Weight Memory};
\node [bottleneck, below=0.45cm of b1] (b2) {KV Cache Memory};
\node [bottleneck, below=0.45cm of b2] (b3) {Attention $\mathcal{O}(n^2)$};
\node [bottleneck, below=0.45cm of b3] (b4) {FFN Bandwidth};
\node [bottleneck, below=0.45cm of b4] (b5) {Sequential Decoding};
\node [bottleneck, below=0.45cm of b5] (b6) {System Throughput};

%% Solution nodes
\node [solution, below=0.5cm of h2]   (s1) {Quantisation (PTQ / QAT)};
\node [solution, below=0.35cm of s1]  (s2) {Pruning};
\node [solution, below=0.35cm of s2]  (s3) {Knowledge Distillation};
\node [solution, below=0.35cm of s3]  (s4) {MQA / GQA};
\node [solution, below=0.35cm of s4]  (s5) {KV Cache Quantisation};
\node [solution, below=0.35cm of s5]  (s6) {PagedAttention};
\node [solution, below=0.35cm of s6]  (s7) {FlashAttention};
\node [solution, below=0.35cm of s7]  (s8) {Speculative Decoding};
\node [solution, below=0.35cm of s8]  (s9) {Continuous Batching};
\node [solution, below=0.35cm of s9]  (s10){llama.cpp / MLC-LLM};

%% Mapping arrows
\draw [arrow] (b1.east) -- (s1.west);
\draw [arrow] (b1.east) -- (s2.west);
\draw [arrow] (b1.east) -- (s3.west);
\draw [arrow] (b2.east) -- (s4.west);
\draw [arrow] (b2.east) -- (s5.west);
\draw [arrow] (b2.east) -- (s6.west);
\draw [arrow] (b3.east) -- (s7.west);
\draw [arrow] (b3.east) -- (s4.west);
\draw [arrow] (b4.east) -- (s1.west);
\draw [arrow] (b5.east) -- (s8.west);
\draw [arrow] (b6.east) -- (s9.west);
\draw [arrow] (b6.east) -- (s10.west);
\end{tikzpicture}
\caption{Mapping of inference bottlenecks (red) to the efficiency techniques (blue)
  surveyed in this chapter. Arrows indicate which technique targets which bottleneck;
  several techniques address multiple bottlenecks simultaneously.}
\label{fig:methods_map}
\end{figure}


\section{Model Compression}\label{sec:compression}

Model compression techniques reduce the size of the model itself---shrinking weight memory and, consequently, the bandwidth required to load those weights during each decode step. They are pre-deployment operations and impose a one-time cost in compute or data for calibration.

\subsection{Quantisation}\label{sec:quantisation}

Quantisation is the process of representing the floating-point weights (and optionally activations) of a trained model at reduced numerical precision. It is currently the most widely adopted compression technique for edge \ac{llm} deployment because it yields large memory reductions with acceptable quality loss and requires no architectural changes.

\paragraph{Why quantisation helps.}
A weight stored in \ac{fp16} occupies 2~bytes. Representing it instead in 8-bit integer format (\ac{int8}) halves weight memory and doubles arithmetic throughput on hardware that natively supports \ac{int8} matrix multiply (e.g., NVIDIA Tensor Cores, ARM Cortex-A78 NEON). Reducing to 4-bit integers (\ac{int4}) achieves a 4$\times$ compression relative to \ac{fp16}, enabling a 7B model to fit in as little as 3.5~GB of memory.

\paragraph{Post-Training Quantisation (\acs{ptq}).}
In \ac{ptq}, a fully trained model is quantised without further gradient-based updates. The simplest approach, round-to-nearest (RTN), maps each weight $w$ to the nearest representable integer value; it is fast but introduces quantisation error that can significantly degrade perplexity below 4 bits.

More sophisticated \ac{ptq} methods achieve substantially better accuracy. \textbf{GPTQ}~\cite{frantar2023gptq} applies a second-order optimisation using the inverse Hessian of the layer-wise reconstruction error to determine quantisation order and compensate remaining weights for the quantisation error of previously quantised weights. This allows accurate weight-only quantisation to \ac{int4} with less than 1~point of perplexity increase on LLaMA-class models.

\textbf{AWQ}~\cite{lin2024awq} takes a complementary perspective: it observes that the importance of a weight is correlated with the magnitude of the corresponding activation, and that protecting a small fraction (1\%) of salient weights in higher precision (while quantising the rest to INT4) preserves most of the model quality. AWQ is activation-\emph{aware} in the sense that it uses a small calibration dataset to identify these salient channels before quantisation.

\textbf{LLM.int8()}~\cite{dettmers2022gpt3int8} addresses quantisation from the activation side. It observes that \ac{llm} activations contain a small number of outlier features with magnitudes 100$\times$ larger than the mean. Quantising these outliers causes disproportionate error. The solution is a mixed-precision decomposition: outlier feature dimensions are kept in \ac{fp16} while the remaining $\sim$99.9\% are multiplied in \ac{int8}, with no loss in model quality at \ac{int8} precision on models above 6.7B parameters.

\paragraph{Quantisation-Aware Training (\acs{qat}).}
In \ac{qat}, the quantisation operation is simulated during training by inserting "fake quantisation" nodes that round activations and weights during the forward pass but compute gradients through the rounding operation via the straight-through estimator. The model therefore learns to be robust to quantisation noise. \ac{qat} typically achieves better accuracy than \ac{ptq} at very low bit widths (2--4 bits), but requires re-training access to the original data and substantial compute---making it less practical when a model's training data is proprietary.

\paragraph{Edge implications.}
Quantisation is the \emph{de facto} entry point for edge \ac{llm} deployment. Tools such as llama.cpp (see Section~\ref{sec:system_level}) implement custom 2--8 bit integer formats (GGUF quantisation) specifically optimised for ARM and x86 CPUs, enabling consumer hardware to run 7B and 13B models interactively. The dominant trade-off is quality vs. compression ratio: below 4 bits, perplexity degradation becomes increasingly noticeable, and the optimal bit-width is task-dependent~\cite{bai2024beyond}.


\subsection{Pruning}\label{sec:pruning}

Pruning removes parameters from the model that contribute least to its predictions, reducing both weight memory and \ac{flop} count.

\paragraph{Unstructured pruning} sets individual weights to zero based on a criterion (most commonly magnitude: $|w_{ij}| < \theta$). High sparsity levels (80--90\%) can be achieved with minimal quality loss for moderately sized models. However, irregular sparsity patterns yield little practical speedup on hardware without sparse compute support, because memory accesses to weight matrices remain irregular and prevent efficient vectorisation.

\paragraph{Structured pruning} removes entire, regular sub-structures---attention heads, \ac{ffn} rows/columns, or entire layers---producing a dense smaller model that achieves genuine speed improvements on standard hardware without requiring specialised sparse kernels.

\textbf{LLM-Pruner}~\cite{ma2023llmpruner} automates structured pruning of \acp{llm} by analysing the dependencies between coupled weight groups (e.g., attention heads and their corresponding projection matrices) and applying gradient-based importance estimation to select which groups to remove. After pruning, a brief post-training or \ac{lora}-based fine-tuning step recovers accuracy. In experiments on LLaMA, LLM-Pruner achieves a 20\% parameter reduction with less than 3~points perplexity degradation and no access to the original pre-training data.

\paragraph{Head pruning} specifically targets attention heads. Studies have shown that many attention heads learn redundant or near-uniform patterns and can be removed without quality loss~\cite{bai2024beyond}. Removing $k$ heads out of $h$ per layer reduces the attention \ac{kv} footprint proportionally, providing a complementary benefit to techniques that reduce \ac{kv} cache size.

\paragraph{Layer pruning} removes entire Transformer blocks. Because blocks are applied sequentially, removing a block reduces latency proportionally. Empirical results on LLaMA-2 show that removing up to 25\% of layers causes less than 5\% perplexity increase on standard benchmarks; the degradation is larger for tasks requiring deep reasoning chains.

\paragraph{Edge implications.}
Structured pruning is particularly attractive for edge deployment because the resulting model is simply a smaller version of the original and requires no specialised runtime. A pruned-then-quantised 7B model can realistically be served on a high-end smartphone: pruning to a 5B effective parameter count followed by INT4 quantisation reduces memory from 14~GB to approximately 2.5~GB.


\subsection{Knowledge Distillation}\label{sec:distillation}

Knowledge distillation (\acs{kd})~\cite{hinton2015distilling} trains a smaller \emph{student} model to mimic the behaviour of a larger \emph{teacher} model. Rather than training the student on hard labels (the ground-truth next token), the student is trained to match the teacher's soft probability distribution over the vocabulary. Because the teacher's output distribution encodes information about semantically related tokens---including near-misses and plausible alternatives---the student receives a richer training signal than hard-label supervision.

\paragraph{Task-specific distillation} fine-tunes a small student on a task-specific corpus using the teacher's outputs as supervision. This produces compact models optimised for narrow applications (e.g., a 1B model for a specific domain), but the student does not generalise beyond the training distribution.

\paragraph{Task-agnostic / general distillation} trains a student from scratch (or from a pre-trained checkpoint) to mimic the pre-training distribution of the teacher. This is far more expensive (requiring access to large corpora and significant compute) but produces a student that retains general-purpose capabilities.

\paragraph{White-box vs. black-box distillation.}
When the teacher's internal representations are accessible (\emph{white-box}), the student can be trained to match not only the output logits but also intermediate hidden states and attention maps, providing additional supervision signal. When only the teacher's output text is available (\emph{black-box}), the student is trained on the teacher-generated text as a dataset.

\paragraph{Edge implications.}
Distillation is the only compression technique that does not start from a large model: it produces a small model from the outset. For edge deployment, 1B and 3B-parameter models distilled from 70B+ teachers (e.g., Phi-2, Phi-3 from Microsoft) have demonstrated that compact architectures can retain a large fraction of the teacher's reasoning capability when the distillation corpus is carefully curated.


\section{Inference-Level Optimisations}\label{sec:inference_opts}

These techniques modify the attention mechanism or the \ac{kv} cache management strategy without altering the model weights. They can be applied at inference time, often transparently to the model.

\subsection{Multi-Query and Grouped-Query Attention}\label{sec:mqa_gqa}

Standard \ac{mha} maintains one independent set of key and value projection matrices per head. With $h$ heads, the \ac{kv} cache therefore stores $h$ distinct key and value tensors per layer and per token, with total per-token cache size $2Lhd_k$.

\textbf{\ac{mqa}}~\cite{shazeer2019mqa} proposes that all query heads share a single key head and a single value head. The \ac{kv} cache therefore collapses to $2L \cdot 1 \cdot d_k$ per token---a factor-of-$h$ reduction. Because the \ac{kv} cache is the second largest contributor to memory consumption after model weights, this reduction is highly significant at edge scale. The trade-off is a slight decrease in model quality (approximately 0.5--1\% on downstream benchmarks), which is generally acceptable~\cite{shazeer2019mqa}.

\textbf{\ac{gqa}}~\cite{ainslie2023gqa} generalises \ac{mqa} by grouping the $h$ query heads into $g$ groups, each sharing one key/value head. With $g = h$ this reduces to standard \ac{mha}; with $g = 1$ to \ac{mqa}. In practice, $g = h/4$ to $h/8$ recovers most of the \ac{mha} quality while approaching \ac{mqa} efficiency. \ac{gqa} has been adopted in LLaMA-2~(70B), LLaMA-3, and Mistral, indicating that the field has largely converged on this as the standard attention variant.

\paragraph{Edge implications.}
Since \ac{mqa}/\ac{gqa} must be incorporated during pre-training, they are not applicable to already-trained models. However, \ac{gqa} can be retrofitted via a brief fine-tuning step (uptraining from \ac{mha} checkpoints) at a fraction of the original training cost~\cite{ainslie2023gqa}. For new models targeting edge deployment, \ac{gqa} should be considered mandatory.


\subsection{FlashAttention}\label{sec:flashattn}

The standard \ac{mha} implementation loads the full $n \times n$ attention score matrix into \ac{hbm} before applying the softmax and value aggregation. For a sequence length of $n = 4096$ tokens with \ac{fp16} precision, this matrix occupies $n^2 \times 2$ bytes $= 32$~MB per head. With 32 heads, this is 1~GB of temporary \ac{hbm} traffic per layer---a dominant bandwidth cost.

\textbf{FlashAttention}~\cite{dao2022flashattention} reformulates the attention computation to avoid materialising the full attention matrix. It tiles the input matrices into blocks that fit entirely in the \ac{gpu}'s fast on-chip \ac{sram}, and fuses the computation of $\mathbf{Q}\mathbf{K}^\top$, softmax, and $\text{softmax}(\cdot)\mathbf{V}$ into a single kernel. This eliminates the round-trip to \ac{hbm} for the intermediate attention matrix, reducing \ac{hbm} accesses from $O(n^2)$ to $O(n)$ while computing the numerically identical result (it is an exact, not approximate, algorithm).

Practically, FlashAttention achieves 2--4$\times$ speedup on attention computation relative to the naive implementation, with proportionally lower energy consumption---making it directly beneficial for edge inference on any hardware with a fast cache hierarchy.


\subsection{KV Cache Optimisation}\label{sec:kvcache_opt}

\paragraph{PagedAttention.}
The \ac{kv} cache grows dynamically with sequence length. In the standard contiguous allocation scheme, each new request pre-allocates a fixed-size maximum-length block of \ac{kv} memory, leading to severe internal fragmentation when requests terminate early. Across thousands of concurrent requests on a server, this wastage can reduce effective throughput by 60--80\%~\cite{kwon2023vllm}.

\textbf{PagedAttention}~\cite{kwon2023vllm}, implemented in the vLLM framework, adapts the virtual memory paging concept to the \ac{kv} cache. Each request's \ac{kv} vectors are stored in fixed-size \emph{pages} that need not be physically contiguous. A page table maps logical to physical page addresses at inference time. This eliminates nearly all internal and external fragmentation, increasing the effective number of concurrent requests that can be served from the same \ac{gpu} memory by 2--4$\times$.

\paragraph{KV cache quantisation.}
The \ac{kv} cache can itself be stored at reduced precision. Quantising the cached key/value tensors to \ac{int8} halves cache memory with minimal perplexity impact. More aggressive quantisation to INT4 is possible with careful calibration, providing a further 2$\times$ reduction.

\paragraph{KV cache eviction.}
For very long sequences, it is possible to \emph{evict} (discard) cache entries for tokens that are unlikely to be attended to by future queries, based on heuristics such as attention score recency or magnitude. StreamingLLM and similar methods maintain a fixed-size cache window, enabling effectively unbounded context length at the cost of some context fidelity.


\section{Decoding Strategies}\label{sec:decoding}

\subsection{Speculative Decoding}\label{sec:speculative}

Standard autoregressive decoding generates one token per forward pass, enforcing a strict latency lower bound of (number of generated tokens) $\times$ (time per forward pass). This is wasteful when the full model is large but most token generations are "easy" (highly predictable).

\textbf{Speculative decoding}~\cite{leviathan2023speculative} exploits this insight with a two-model approach:
\begin{enumerate}
    \item A small, fast \emph{draft model} (e.g., 1--3B parameters) generates $K$ candidate tokens autoregressively in $K$ forward passes.
    \item The large \emph{target model} processes all $K$ draft tokens in a single forward pass (which is parallelisable, like prefill) and accepts or rejects each draft token using a rejection sampling criterion that guarantees the accepted sequence has the same distribution as the target model's output.
\end{enumerate}
When the draft model's predictions match the target with probability $\alpha$ (the \emph{acceptance rate}), the expected number of tokens produced per target-model forward pass is $\frac{1 - \alpha^{K+1}}{1-\alpha}$. For $\alpha \approx 0.8$ and $K = 5$, this yields approximately 3.5 tokens per call to the target model---a $3.5\times$ latency improvement with no loss in output quality~\cite{leviathan2023speculative}.

For edge deployment, the "large model" and "draft model" may themselves be a quantised medium model and a tiny model respectively. The framework is flexible: any two models with matching tokenisers and vocabulary can be paired.

\subsection{Continuous Batching}\label{sec:batching}

Traditional static batching waits until a fixed number of requests arrive and then processes them together. Requests of different lengths must be padded to a common length (wasting compute) and the entire batch must finish before new requests are admitted (increasing average latency).

Continuous (or iteration-level) batching inserts new requests into the running batch at the earliest available slot (e.g., whenever a request produces an EOS token and vacates its slot). This keeps hardware utilisation high and reduces median latency significantly. Combined with PagedAttention, continuous batching enables high-throughput serving that is relevant for cloud deployments and, increasingly, for shared edge inference servers in enterprise settings.


\section{System-Level Runtimes and Edge Frameworks}\label{sec:system_level}

Efficiency techniques are only realised in practice when they are implemented in well-engineered software runtimes that expose hardware capabilities efficiently. Several frameworks have emerged specifically for \ac{llm} inference, with differing strengths on server and edge hardware.

\subsection{llama.cpp}

\texttt{llama.cpp}~\cite{ggerganov2023llamacpp} is a C/C++ inference runtime designed for deployment on commodity hardware, including ARM CPUs (Apple Silicon, Android), x86 CPUs, and consumer \acp{gpu}. Its key design decisions include:
\begin{itemize}
    \item \textbf{GGUF quantisation format}: a family of mixed-precision weight quantisation schemes (2--8 bits per weight) with hand-tuned dequantisation kernels for NEON (ARM SIMD), AVX2, and CUDA.
    \item \textbf{Minimal dependencies}: the library has no deep learning framework dependency, making it straightforward to compile and deploy on embedded targets.
    \item \textbf{Metal / Vulkan backend}: hardware-accelerated compute on Apple Neural Engine and Mali/Adreno \acp{gpu} on mobile devices.
\end{itemize}
In practice, llama.cpp enables a quantised (Q4\_K\_M) 7B model to generate tokens at 20--40 tokens/second on a modern laptop CPU and at 60--80 tokens/second on an Apple M-series chip---sufficient for interactive use~\cite{edgellm2024review}.

\subsection{MLC-LLM}

\textbf{MLC-LLM}~\cite{chen2023mlc} (\ac{mlc} Large Language Model) takes a compilation-based approach. It uses the Apache TVM deep learning compiler to generate hardware-specific compute kernels from a high-level model description, targeting iOS (Metal), Android (OpenCL, Vulkan), WebGPU, CUDA, and ROCm from a single code base. Key advantages include:
\begin{itemize}
    \item Auto-tuning of tiling and scheduling parameters for the target hardware.
    \item Support for both weight quantisation and \ac{kv} cache quantisation.
    \item On-device inference on smartphones (e.g., LLaMA-2 7B at 3--8 tokens/second on a Pixel 7 Pro).
\end{itemize}
The compilation step adds several minutes of one-time overhead but produces deployment artefacts that run efficiently with no runtime compilation cost.

\subsection{vLLM}

\textbf{vLLM}~\cite{kwon2023vllm} is optimised for \emph{server-side} high-throughput serving. Its core contribution is PagedAttention (Section~\ref{sec:kvcache_opt}), which dramatically reduces \ac{kv} cache memory fragmentation. vLLM additionally implements continuous batching, FlashAttention, and quantised inference, providing a production-grade serving stack for data-centre deployment. While not directly targeted at edge hardware, vLLM establishes the baseline for cloud-side efficient inference from which edge-focused frameworks draw inspiration.

\subsection{Comparison and Edge Suitability}

Table~\ref{tab:framework_comparison} compares the three frameworks across the dimensions most relevant to edge deployment.

\begin{table}[ht]
\centering
\caption{Comparison of selected inference frameworks for edge deployment.
  \checkmark\ = supported, $\sim$\ = partial support, $\times$\ = not supported.}
\label{tab:framework_comparison}
\small
\begin{tabular}{lcccc}
  \hline\hline
  \textbf{Framework} & \textbf{Edge HW} & \textbf{Mobile} & \textbf{Quantisation} & \textbf{Throughput focus} \\
  \hline
  llama.cpp         & \checkmark & \checkmark & \checkmark (2--8 bit) & \checkmark \\
  MLC-LLM           & \checkmark & \checkmark & \checkmark (4--8 bit) & $\sim$ \\
  vLLM              & $\times$   & $\times$   & $\sim$ (INT8)          & \checkmark \\
  \hline\hline
\end{tabular}
\end{table}


\section{Discussion: Trade-offs and Edge-Specific Considerations}\label{sec:tradeoffs}

No single technique dominates across all deployment scenarios. The efficiency gains and quality losses described above interact in ways that must be understood holistically when targeting edge hardware.

\subsection{The Efficiency--Accuracy Trade-off}

Compression techniques invariably introduce some quality degradation. The magnitude of this degradation depends on:
\begin{itemize}
    \item \textbf{Bit width}: quantisation from FP16 to INT8 is nearly lossless; from INT8 to INT4 introduces a noticeable but usually acceptable quality reduction; below INT4, degradation accelerates~\cite{frantar2023gptq}.
    \item \textbf{Task sensitivity}: arithmetic, logical reasoning, and mathematical tasks degrade faster under quantisation than open-ended text generation~\cite{bai2024beyond}.
    \item \textbf{Model scale}: larger models exhibit greater robustness to compression. A 70B model quantised to INT4 often outperforms an uncompressed 7B model, making the question "which compressed large model vs. which uncompressed small model" a central practical question for edge deployment.
\end{itemize}

\subsection{Orthogonality and Composability}

Many of the techniques described in this chapter are \emph{orthogonal}: they address different bottlenecks and can be applied simultaneously. A practical edge stack might combine:
\begin{enumerate}
    \item \textbf{Distillation} (at training time) to obtain a smaller base model.
    \item \textbf{\ac{gqa}} (in the model architecture) to reduce \ac{kv} cache size.
    \item \textbf{GPTQ or AWQ quantisation} (post-training) to compress weights to INT4.
    \item \textbf{FlashAttention} (at inference time) to reduce attention bandwidth.
    \item \textbf{Speculative decoding} (at inference time) to improve throughput.
    \item \textbf{llama.cpp or MLC-LLM} as the runtime to execute efficiently on ARM/mobile hardware.
\end{enumerate}
Applied together, these techniques have enabled 7B-class models to be deployed interactively on smartphones in 2024---an outcome that would have seemed impractical just two years earlier.

\subsection{The Fundamental Edge Constraint: Memory}

Across all dimensions, memory remains the hardest constraint on edge devices. Compute and energy can often be traded against latency by reducing clock frequency; bandwidth can be partially mitigated by caching. But there is no recourse when a model simply does not fit in the available \ac{dram}: inference is not possible. This makes weight quantisation and architectural choices that reduce the \ac{kv} cache (such as \ac{gqa}) the most impactful interventions for edge deployment, not as performance optimisations but as enabling technologies~\cite{edgellm2024review}.
