% !TEX root = thesis.tex
%******************************************************************************
% Assurance everything is single handed
%******************************************************************************

\begingroup
\thispagestyle{empty}
\begin{center}
	\large
	\textbf{Eidesstattliche Versicherung}
\end{center}

\vspace{0.5cm}

\begin{tabular}{@{}p{8cm}p{5.8cm}}
	\underline{\hspace{6cm}} & \underline{\hspace{5.8cm}} \\
	\vspace{0.02cm}Name, Vorname & \vspace{0.02cm}Matrikelnummer \\
\end{tabular}
\vspace{0.5cm}

Ich versichere hiermit an Eides Statt, dass ich die vorliegende {\thesistypegerman} mit dem Titel

{\makeatletter{\emph{{\@title}}}\makeatother}

selbständig und ohne unzulässige fremde Hilfe erbracht habe. Ich habe keine anderen als die angegebenen Quellen und Hilfsmittel benutzt. Für den Fall, dass die Arbeit zusätzlich auf einem Datenträger eingereicht wird, erkläre ich, dass die schriftliche und die elektronische Form vollständig übereinstimmen. Die Arbeit hat in gleicher oder ähnlicher Form noch keiner Prüfungsbehörde vorgelegen.

\vspace{0.3cm}

\begin{tabular}{@{}p{8cm}p{5.8cm}}
	\underline{\smash{Aachen, den \today}} & \underline{\hspace{5.8cm}}\\
	\vspace{0.02cm}Ort, Datum & \vspace{0.02cm}Unterschrift \\
\end{tabular}

\vspace{1cm}

\begin{small}
	\textbf{Belehrung:}

	\textbf{§ 156 StGB: Falsche Versicherung an Eides Statt}\\
	Wer vor einer zur Abnahme einer Versicherung an Eides Statt zuständigen Behörde eine solche Versicherung falsch abgibt oder unter Berufung auf eine solche Versicherung falsch aussagt, wird mit Freiheitsstrafe bis zu drei Jahren oder mit Geldstrafe bestraft.

	\textbf{§ 161 StGB: Fahrlässiger Falscheid; fahrlässige falsche Versicherung an Eides Statt}\\
	(1) Wenn eine der in den §§ 154 bis 156 bezeichneten Handlungen aus Fahrlässigkeit begangen worden ist, so tritt Freiheitsstrafe bis zu einem Jahr oder Geldstrafe ein.\\
	(2) Straflosigkeit tritt ein, wenn der Täter die falsche Angabe rechtzeitig berichtigt. Die Vorschriften des § 158 Abs. 2 und 3 gelten entsprechend
\end{small}

\vspace{0.3cm}

Die vorstehende Belehrung habe ich zur Kenntnis genommen:

\vspace{0.3cm}

\begin{tabular}{@{}p{8cm}p{5.8cm}}
	\underline{\smash{Aachen, den \today}} & \underline{\hspace{5.8cm}}\\
	\vspace{0.02cm}Ort, Datum & \vspace{0.02cm}Unterschrift \\
\end{tabular}

\endgroup
\cleardoublepage

%******************************************************************************
% Abstract
%******************************************************************************

\begingroup
\begin{abstract}
\thispagestyle{plain}
Large generative models, in particular decoder-only Transformer architectures, have demonstrated unprecedented capabilities in natural language processing, code synthesis, and multimodal reasoning. However, the inference process is computationally intensive: repeated autoregressive forward passes impose severe demands on memory, compute, memory bandwidth, and energy. These characteristics make deployment on resource-constrained edge devices --- such as smartphones, embedded systems, and \ac{iot} hardware --- a significant open challenge.

This paper provides a structured survey of resource-efficient inference in large generative models. Starting from the fundamental Transformer building blocks, it traces the complete inference pipeline and systematically identifies the principal resource bottlenecks at each stage, distinguishing between the compute-bound prefill phase and the memory-bandwidth-bound decode phase. Building on this analysis, the paper surveys the leading efficiency techniques across four layers: model compression (quantisation, pruning, and knowledge distillation), inference-level algorithmic improvements (optimised attention mechanisms and \ac{kv} cache management), decoding strategies (speculative decoding and batching), and system-level runtimes (vLLM, llama.cpp, and \acs{mlc}-LLM). For each technique, the discussion focuses on the specific bottleneck it addresses and the practical trade-offs that arise when targeting edge hardware.

The analysis shows that quantisation combined with grouped-query attention and speculative decoding currently represents the most promising combination for practical edge inference, enabling models with billions of parameters to run on consumer-grade hardware with acceptable quality degradation.

\end{abstract}
\endgroup
\cleardoublepage

%******************************************************************************
% TOC
%******************************************************************************

\tableofcontents
\cleardoublepage

%******************************************************************************
% List of figures and tables
%******************************************************************************

\listoffigures
\cleardoublepage

\listoftables
\cleardoublepage

%******************************************************************************
% List of abbreviations
%******************************************************************************

\printglossary[type=\acronymtype, title={List of Abbreviations}, style=list]
\cleardoublepage
